\vspace{1.0cm} \LARGE \bf Supporting Text: \\ Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2013-01-22 flux_si.tex
Fix typo.
Confirm that diffusion coefficient estimator has lower mean squared error than estimator used in Pybus et al 2012 with some estimation error on node locations and branch lengths.

The distance $d$ traveled by two-dimensional diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
\begin{equation}
	X_t \sim \normal(0, \sigma^2 t),
\end{equation}
where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
\begin{equation}
	\mathrm{E}[X_t^2] = \sigma^2 t.
\end{equation}
In moving to a two-dimensional diffusion
\begin{equation}
	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma^2 t}{0}{0}{\sigma^2 t} \right),
\end{equation}
a factor of 2 is included when calculating expected squared displacement
\begin{equation}
	\mathrm{E}[X_t^2 + Y_t^2] = 2 \sigma^2 t.
\end{equation}
Such diffusions are often parameterized with diffusion coefficient $D$ rather than volatility parameter $\sigma$, in this case $D=\sigma^2/2$ and 
\begin{equation}
	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
\end{equation}

Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
\begin{equation} \label{pybusest}
	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
\end{equation}
in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
However, short branches along the phylogeny may cause $d_i^2/t_i$ to have high variance and lead to uncertainty in the resulting estimates of $D$.
Consequently, we estimate $D$ by looking at the relationship between total squared displacement along branches of a phylogeny and total time along branches of a phylogeny
\begin{eqnarray}
	d^2_\mathrm{total} = \sum^n_{i=1} d^2_i \nonumber \\
	t_\mathrm{total} = \sum^n_{i=1} t_i^2 \nonumber \\
	\hat{D} = \frac{ d^2_\mathrm{total} }{ 4 t_\mathrm{total} }.
\end{eqnarray}
This estimator showed less variance than the estimator in equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
When estimating $D$ specific to trunk branches or side branches, we calculate $d^2_\mathrm{total}$ and $t_\mathrm{total}$ only from the subset of trunk branches or side branches as appropriate.

>>>>

The distance $d$ traveled by two-dimensional diffusion is not a linear function of time interval $t$, and thus the `rate' of diffusion cannot be calculated following $r=d/t$.
Assume we have a one-dimensional diffusion starting from $x_0 = 0$, so that
\begin{equation}
	X_t \sim \normal(0, \sigma^2 t),
\end{equation}
where $\sigma$ represents the diffusion volatility, and in this case, the expected squared displacement follows
\begin{equation}
	\mathrm{E}[X_t^2] = \sigma^2 t.
\end{equation}
In moving to a two-dimensional diffusion
\begin{equation}
	(X_t, Y_t) \sim \normal \left( \twomatrix{0}{0}, \fourmatrix{\sigma^2 t}{0}{0}{\sigma^2 t} \right),
\end{equation}
a factor of 2 is included when calculating expected squared displacement
\begin{equation}
	\mathrm{E}[X_t^2 + Y_t^2] = 2 \sigma^2 t.
\end{equation}
Such diffusions are often parameterized with diffusion coefficient $D$ rather than volatility parameter $\sigma$.
In this case $D=\sigma^2/2$ and 
\begin{equation}
	\mathrm{E}[X_t^2 + Y_t^2] = 4 D t.
\end{equation}

Owing to this relationship, Pybus et al.\ \cite{Pybus12} estimate $D$ from phylogenies in which internal nodes have inferred 2D locations via the formula
\begin{equation} \label{pybusest}
	\hat{D} = \frac{1}{n} \sum^n_{i=1} \frac{d_i^2}{4t_i},
\end{equation}
in which $d_i$ represents the displacement of a branch and $t_i$ represents the length of a branch.
However, uncertainty in $d_i$ or $t_i$, especially along short branches of the phylogeny, may cause $d_i^2/4t_i$ to have high variance and increase uncertainty in the resulting estimates of $D$.
Consequently, we estimate $D$ by looking at the relationship between total squared displacement and total time along branches of a phylogeny
\begin{eqnarray} \label{ourest}
	d^2_\Sigma &=& \sum^n_{i=1} d^2_i \nonumber \\
	t_\Sigma &=& \sum^n_{i=1} t_i \nonumber \\
	\hat{D} &=& \frac{ d^2_\Sigma }{ 4 t_\Sigma }.
\end{eqnarray}
This estimator showed less variance than equation \ref{pybusest} when analyzing tree output from the BMDS model and was thus preferred.
Additionally, we confirmed through simple simulation that estimator \ref{ourest} shows lower mean squared error than estimator \ref{pybusest} in the presence of modest levels of noise on estimates of $d_i$ and $t_i$.
When estimating $D$ specific to trunk branches or specific to side branches, we calculate $d^2_\Sigma$ and $t_\Sigma$ only from a subset of phylogeny branches as appropriate.
